Using structural and evolutionary information to detect and correct pyrosequencing errors in non-coding RNAs

### 2013-05-28 JCB/ResponseToReviewers-RNAPyro-JCB-2013.tex

\begin{document}
\maketitle

In general, the reviewers did not raise very specific points, but were mostly concerned by the applicability of the method in a high-throughput context. In this extended version, we demonstrated the scalability of the method by running processing longer sequences (16S rRNAs) obtained from random reads (generated by the {\tt ART} pyrosequencing simulator), a setting which is closer to real use-cases in metagenomics. Furthermore, our python implementation  is still more time-consuming than would be ideal in a high-throughput context. However, we propose a pipeline which first maps short reads to the reference alignment, followed by an alignment/correction of the consensus sequence against the reference multiple sequence alignment. In this setting, {\tt RNAPyo} is only executed once for each rRNA, as opposed to once for every read, and the typical runtime of our software does not induce such a strong overhead. Additionally, we investigated the error-correction potential as a function of the coverage (depth) of the sequencing.

\section*{Detailed responses}

\Comment{(Reviewer \#1) While the paper is algorithmic interesting, and clearly publishable in principle, I'm a bit concerned with the practical application of RNApyro. If many reads are available for the same sequence, the tool is obviously not necessary since purely sequence-based error detection should be applicable. In the scenario envisioned here, however, the model makes a very strong assumption, name a completely known consensus structure to which the query sequence must fit without indels. While plausible if the consensus is derived from closely related species (in which case point mutations that disrupt the structure are easily visible in most cases) one would expect also structural variations as well.\\
(Reviewer \#2) However, assuming that it works as expected, I understand that a successful "error correction" involves replacing an unlikely secondary structure base pair by a more likely one. Couldn't this become a problem for outlier RNA sequences that do not exactly fold according to the canonical model used in RFAM? If this is the case, then the program would correct certain sequences/species better than others, which is not quite a desirable property in error correction. 
}
\Answer{We agree with the reviewer that the error-correction potential of our method may be challenged by sequences whose folding is substantially different from the consensus structure. However, assuming some level of structural conservation within RNA families that share a given function is generally held as reasonable, and constitutes the main paradigm in RNA comparative bioinformatics, e.g. underlying the RFAM database and homologous sequences retrieval. This paradigm is, of course, challenged in the case of structure-disrupting mutations (which are increasingly identified and linked with RNA-based genetic conditions), but should remain reasonably consensual in the context of metagenomics, where structural outliers (often pathological, or even lethal) are perhaps of lesser interest. \\
Furthermore, the consensus structure needs not be completely known. In this extended version, we considered the potential for error-correction of {\tt RNAPyro}, restricted paired regions, based on the dual rationale that 1) our energy model does not (yet!) capture unpaired regions, and 2) unpaired region may be paired but not sufficiently conserved throughout the evolution (or simply misaligned). The excellent areas under the ROC curves observed in this case suggests that a partial consensus may not too strongly impede the quality of our error corrections. \\
Finally, regarding a more general structural diversity in families of structural RNAs (e.g. only preserving a domain, a opposed to a general architecture), we believe such cases would be revealed by the level of detail of the consensus structure and, by disallowing error-corrections within unpaired regions, our method should not be overly erroneous and inhomogeneous (at the obvious cost of missing corrections within unpaired regions). }

### 2013-05-28 JCB/results_JCB.tex
comment time & URL
method

The time benchmarks were performed on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM.
Since typical use-cases of \RNApyro require
 efficiency and scalability, we present in Table~\ref{tab:time} typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those tests,
 both the sequences and the target secondary structure were generated at random.
\TODOVlad{Add more extensive time benchmark, as a plot}

@@

\subsection{Error Correction in 5s rRNAs}

To illustrate the potential of our algorithm, we applied our techniques to identify and correct point-wise errors in RNA sequences
with conserved secondary structures. More precisely, we used \RNApyro to reconstruct 5s rRNA sequences with randomly distributed
mutations. This experiment has been designed to suggest further applications to error-corrections in pyrosequencing data.

@@

>>>>

The time benchmarks were performed on a MacMini 2010, 2.3GHz dual-core Intel Core i5, 8GB of RAM for the Sec.~\ref{sec:5S} and Sec.~\ref{sec:16S} on AMD Opteron(tm) Processor 6278  at 2.4 GHz with cache of 512 KB.20 cores and 175 GB ram. 
Since typical use-cases of \RNApyro require
 efficiency and scalability, we present in Table~\ref{tab:time} typical runtimes required to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those tests,
 both the sequences and the target secondary structure were generated at random.
\TODOVlad{Add more extensive time benchmark, as a plot}

@@

\subsection{Error Correction in 5s rRNAs}
\label{sec:5S}
To illustrate the potential of our algorithm, we applied our techniques to identify and correct point-wise errors in RNA sequences
with conserved secondary structures. More precisely, we used \RNApyro to reconstruct 5s rRNA sequences with randomly distributed
mutations. This experiment has been designed to suggest further applications to error-corrections in pyrosequencing data.

@@

\subsection{Error Correction in 16s rRNAs}
\label{sec:16S}

To complete our benchmark we also looked at the Bacterial small subunit ribosomal RNA. 
We gathered the seed multiple sequence alignment (MSA) composed of 93 sequences from the Rfam Database 11.0 (Rfam id: RF00177)~\cite{gardner2011rfam}.
Those sequences are between 1461 and 1568 nucleotides long and yield a similarity of $69\%$.

As in Sec.~\ref{sec:5S} we used a leave-one-out strategy. Since few structures contain all gaps 
at the same positions, after selecting a sequence of interest,  we used the following strategy.
 It was applied to each sequence in the alignment. Starting with the full alignment, all positions corresponding to gaps in the sequence of interest were removed, for all sequences in the alignment.

To simulate sequencing error in the sequence of interest,
 we use the the next-generation sequencing read simulator 
ART~\cite{huang2012art}. The Illumina technology main error mode is base substitution. We used
this model since those
are the errors that \RNApyro detects. As parameters, the reads reads generated
were $75\text{bps}$ long for an average fold coverage of $5$, $10$ and $15$. 
 They yield respectively an average error rate of $2.4\%$, $0.9\%$ and $0.6\%$.

 As in Sec.~\ref{sec:5S} we used a value of $\beta=15$ and evaluated $\alpha$ at $0, 0.5, 0.8, 1.0$. 
 Additionally since the alignment still contains gap, we made sure to evaluate the isostericity term
